---
title: "Lab 2 - Data Management for Housing Analysis"
author: |
  | **Esteban Lopez Ochoa**
  | *URP 5313 - Urban Housing Policy and Analysis*
  | University of Texas at San Antonio
date: "10/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The objective of this lab is to provide an overview of the main methods to import, manipulate and analyze data related to housing issues. This lab is organized in three parts:

- Part 1: Using & manipulating data from an external source
- Part 2: Indentifying key housing issues and data gaps

BY the end of this lab, the student is expected to have a broad knowledge and basic hands on experience of the methods to import and analyse data relating to housing issues.


### Part 1: Using data from an external source

Data is comonly first downloaded from an external source and then loaded into R for procesing. This script shows how to load data from an external source and how to process it in order to create outputs (i.e. maps, tables, etc.) that will serve to inform policymaking.

###### Loading an external Data Source: The Opportunity Insights Project

Opportunity insights is a not-for-profit organization directed by Prof. Raj Chetty (Harvard University) developed with the aim to **"...identify barriers to economic opportunity..."**. As housing scholars, we know that for a increasingly larger portion of the population, housing is a steping stone over which famkilies can build wealth and create stability and opportunities for their families. Hence much of the barriers to economic opportunities identified in the OI project overlap with the barriers for attaining a health housing situation. In specific the neighborhood dimension of the OI project contains many important findings that can provide useful insights when embarking in a housing analysis journey (see more in: https://opportunityinsights.org/neighborhoods/)

For our excercise, we will use the neighborhood variables at the census tract level.This data set can be downloaded manually from:https://opportunityinsights.org/data/, by selecting `census tract` geographic level and then downloading the excel file for 'Neighborhood characteris by Census Tracts'. Alternativelly, the data can be directedly downloaded and imported directly using the following:

```{r OI Data Import}
library(data.table)

d1<-fread("https://opportunityinsights.org/wp-content/uploads/2018/10/tract_covariates.csv")
d2<-fread("tract_covariates.csv")

identical(d1,d2)
rm(d2)
```

Always it is useful to have a good idea of the strcuture of the data, namely, what variables are there available, what format are they stored, and what limitations (i.e. missing, incomplete cases, etc.) the data might have. This can be achieve by: 

1. Looking at the data source metadata. In the case of the OI data this information can be found at: https://opportunityinsights.org/wp-content/uploads/2019/07/Codebook-for-Table-9.pdf 
2. Looking at the structure of the imported data and corroborating if all variables were imported correctinly, this can be done by:


```{r OI data structure}
str(d1) # provides a general structure of the data (variable names, type, forst 6 observations)

summary(d1) # provides a detailed summary statistic by variable (inlcuding the presence of NA's)

class(d1)

```

###### Most common data manipulation operations

```{r OI data manipulation}

d1[,table(czname)] # variable exploration - counts
d1[,.N, by=.(czname)] #alternative

d1[czname=="San Antonio",] # filtering

SA_OI<- d1[czname=="San Antonio",] # filtering + object creation (subsetting)

SA_OI[is.na(hhinc_mean2000),] #identifying cases with NAs

SA_OI<- SA_OI[is.na(hhinc_mean2000)==FALSE,]

SA_OI[, ones:=1] # variable creation
SA_OI[hhinc_mean2000<=quantile(hhinc_mean2000,probs = 0.5), bellow_medianIncome:=1 ] # variable creation by codition
SA_OI[is.na(bellow_medianIncome),bellow_medianIncome:=0] # variable value replacement by condition

SA_OI[, med_hhinc_growth1990_2006:=((med_hhinc2016-med_hhinc1990)/med_hhinc1990)*100]# variable creation by a mathematical manipulation of other variables.

```


###### More advanced data manipulation

```{r IO advanced data manipulation}

#let's load a simple map
library(tidycensus)

census_api_key("0d539976d5203a96fa55bbf4421110d4b3db3648")# you must acquired your own key at http://api.census.gov/data/key_signup.html

bexar_medincome <- get_acs(geography = "tract", variables = "B19013_001",
                state = "TX", county = "Bexar", geometry = TRUE)

bexar_medincome

plot(bexar_medincome)


#merge SA_OI data with the map

head(SA_OI) ; head(bexar_medincome)

SA_OI[,GEOID:=paste0(state,"0",county,tract)] #creating a GEOID variable to have a common variable between the two data sets

#check variable classes

class(bexar_medincome$GEOID)
class(SA_OI$GEOID)

table(bexar_medincome$GEOID %in%  SA_OI$GEOID) # checking overlap


# merge/join by GEOIDS
bexar_medincome2<-merge(bexar_medincome,SA_OI,by="GEOID")


#plotting

plot(bexar_medincome2[,"med_hhinc_growth1990_2006"])# fast plotting

library(ggplot2);library(viridis) # prettier plotting
ggplot(bexar_medincome2)+
  geom_sf(aes(fill=med_hhinc_growth1990_2006))+
  scale_fill_viridis(option = "magma") +
  scale_color_viridis(option = "magma")

library(leaflet) # dynamic plotting

pal <- colorQuantile("YlOrRd", domain = bexar_medincome2$med_hhinc_growth1990_2006, n = 5)

leaflet(bexar_medincome2)%>%
  addProviderTiles(provider = providers$CartoDB.Positron)%>%
  addPolygons(fillColor = ~pal(med_hhinc_growth1990_2006),label = ~med_hhinc_growth1990_2006,color=~pal(med_hhinc_growth1990_2006),fillOpacity = 0.5,weight = 0.1)%>%
  addLegend(pal = pal,values = ~med_hhinc_growth1990_2006,opacity = 1)

```



### Part 2: Indentifying key housing issues and data gaps 

###### Exploratory data analysis

The objective of an Exploratory Data Analysis (EDA) is to explore and reveal relationships between variables available in the data as a way to provide initial insights on how a particular dataset can be useful to inform the researcher about a particular problem. 

In housing analysis, we know that there are two main types of variables that we may be able to acquire data about:

  1. Housing sympthoms variables: Variables that provide information about a sympthom or issue directly related to the health status of a housing market. Examples of these variables are: 
    - Affordability Index: assess the levels of affordability in a particular area (can be defined in many ways) - housing demand
    - New construction starts: assess the levels of dynamism of housing supply
    - Housing prices: asssess the level of housing demand
    - Housing tenure rates: assess the status of housing demand between the sales and renting market
    - Eviction/ foreclosure counts: assess the level of how critical is a household' housing situation 
  
  2.Housing exogenous forces: variables that provide information about socio-economic forces that affect the housing market, but are normally contextual and outside of the control of planning/housing actors. Examples of these variables are:
    - Employment/Unemployment: Variables related to job creation or destruction.
    - Private investment: variables related the private companies investing in land aquisition or job creation
    - Interest rates: indicator of how attractive can be the mortgage sector
    - Mortgage placements: variables related to the lending market
    - Insurance variables: variables related to how the insurance market is covering housing market mortgages/rental leases
    - Climate related variables: variables related to climate events affecting the housing market (e.g., flooding maps, etc.)
    - Demographic variables: variables that alone don't tell much about housing, but need to be spatially referenced or combined with other housing variables to do so.


_Excercise 1_: Let's create two objects listing the variables according to the to previous groups. Then create a `data.table` object for each of the variable types.     

```{r OI Variable filtering, echo=FALSE}
names(d1)


```


_Excercise 2_: Alternativelly, in the case we want to keep all variables in the same object, how can we identify them?      

```{r OI Variable rename, echo=FALSE}
names(d1)


```


Now that we have classified the variables by their type, we can start exploring what relations could be present in the data. This is a recursive process that entails going over the following questions:

  1. What problem I would like to seek/visualize in the data?
  2. What is the hipothesis behind? 
  3. What variables do I have in my data to visualize that hipothesis?
  4. Are the variables in the right format? or do I need to do data manipulation?
  5. What is the best way to display the data to inform about the problem, issue?
  6. If the resulting product (table/map/figure/graph/etc.) informative of the issue? 
  7. Go back to question 1 and 2 and see if you achieved the objective.



##### Housing issue 1: Urban segregation




##### Housing issue 2: 
